\section*{Acknowledgment}

This project benefited from substantial contributions from Mike Wiser, who co-authored earlier work related to this project \citep{lalejini2017gene}.
We also would like to acknowledge Jeffrey E. Barrick and David M. Bryson for laying the groundwork for this research.
Additional thanks to Rosangela Canino-Koning and Emily Dolson for their comments and the Digital Evolution Laboratory for thoughtful discussions, ideas and support.

This research was supported in part by the BEACON Center for the Study of Evolution in Action and by Michigan State University through the computational resources provided by the Institute for Cyber-Enabled Research.

This research was supported in part by the NSF Grant DEB 1813069.

This material is based upon work supported by the Eric and Wendy Schmidt AI in Science Postdoctoral Fellowship, a Schmidt Sciences program.

This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research (ASCR), under Award Number DE-SC0025634.
This report was prepared as an account of work sponsored by an agency of the United States Government.
Neither the United States Government nor any agency thereof, nor any of their employees, makes any warranty, express or implied, or assumes any legal liability or responsibility for the accuracy, completeness, or usefulness of any information, apparatus, product, or process disclosed, or represents that its use would not infringe privately owned rights.
Reference herein to any specific commercial product, process, or service by trade name, trademark, manufacturer, or otherwise does not necessarily constitute or imply its endorsement, recommendation, or favoring by the United States Government or any agency thereof.
The views and opinions of authors expressed herein do not necessarily state or reflect those of the United States Government or any agency thereof.
