\section{Introduction} \label{sec:introduction}

A key objective of evolutionary biology is to understand how evolution gives rise to new genetic information and complex phenotypic traits, and gene duplication is one prime suspect \citep{Zhang2003,Otto2000,Wagner2008,Crow:2006role,Magadum:2013wu,Metz:chromosomeDuplication1947,Hu:2010ea,Castellanos2024,Lynch2000b,Lynch2003,Birchler2022}.
While comparative studies provide increasingly detailed accounts of duplication events shaping specific evolved traits \citep{Zhang2014,teichmann_structural_1998,Teichmann:2004cz,DuBose2024,Voordeckers2012,Pougach2014,Moore2003,Ogino2015,Tasnim2024,Assis2013,Deng2010,Monson2003,Xue2026,Lynch2007Inventing}, work remains to develop systematic understanding \citep{Welch2016} of gene duplication's broader role in both adaptive and neutral evolutionary processes \citep{Innan2010,Zhang:2003fw,Kuzmin2022}.
In this paper, we use direct experimentation to test how gene duplication impacts structural and functional complexity, and the extent to which such impact depends on the actual genetic information content within duplications.

% Mirroring larger questions around the adaptive framings of evolutionary origins of biological complexity, gene duplication is thought to lead to increased complexity both by increasing opportunities for the accumulation of contingent complexity (``sub-functionalization'') and by facilitating adaptive discovery of beneficial novel traits (``neo-functionalization'').
In recent years, such experimental approaches --- which can explicitly test causality --- have become an important complement to observational studies of natural history \citep{Kawecki2012,Elena2003}.
In the context of gene duplication, microbial experiments have confirmed the adaptive significance of duplicative dose effects \citep{Tong2025,Dhar2014,Nsvall2012}.
Another notable microbial experiment has tested how gene copy count affects evolutionary innovation \citet{mihajlovic2025direct}.
Comparing evolution of fluorescence phenotypes between single- and double-copy coGFP populations, the latter populations exhibited greater robustness under mutation but no faster adaptive change under selection.
As the authors highlight, though, practical constraints required they impose mutation rates and selection pressures far removed from natural evolution.

More broadly, experimental work has also made notable headway in exploring how evolution can produce complexity.
Here, key contributions have relied on digital model systems \citep{Fortuna2022,Adami2000Evolution,Zaman2014Coevolution,Fortuna2017GenotypePhenotype}.
Though substantially abstracting mechanistic aspects of biological organisms \citep{peck2004simulation}, such \textit{in silico} experiments can bring within reach principled definitions of complexity \citep{Adami2002}, fast throughput spanning thousands of generations \citep{moreno2024trackable}, and flexible exploration of arbitrary counterfactuals \citep{langton1989artificial}.
In this work, we rely on digital capabilities to impose wholesale suppression of duplicative mutations, as well as intermediate interventions partially disabling them.
We also exploit the opportunity to trace duplicated sites across exact lineage histories \citep{dolson2024phylotrack}, and thereby directly ascertain their role in later innovation.

\input{fig/slip_mut_variants.tex}

We conducted digital experiments using Avida, software which administers populations of self-replicating computer programs --- providing, essentially, a digital ``model organism'' \citep{Ofria:2009avida}.
Because copy errors introduce heritable variation in how efficiently programs replicate, evolution by natural selection results \citep{pennock2007models}.

In early work with Avida, Lenski \textit{et al.} established a useful framework for studying phenotypic traits across a well-defined spectrum of complexity;
input/output calculations can be treated as phenotypic traits, with the complexity of a given trait given according to its minimum necessary substeps.
Applying this lens, Lenski \textit{et al.} demonstrated that simpler traits could provide building blocks for complex ones --- and, in fact, often amounted to necessary prerequisites for their evolution \citep{Lenski2003Evolutionary}.

To complement the above phenotypic measure of functional complexity, we also established a genetic measure of structural complexity.
Here, we counted genome instructions that benefit an organism's fitness \citep{Adami2002}.
These adaptive sites were identified through single-site knockout screens, disabling genome instructions one-by-one via null placeholders.
\unskip\footnote{%
Although sufficiently representative for our purposes, limitations exist in detecting Avida genome functionality through single-site knockouts; such an approach can underestimate aspects of genome sequence complexity involving small effects or redundancy \citep{Lenski1999,moreno2024cryptic}.
}

With respect to both our functional and structural metrics, we find that gene duplication can systematically increase evolved complexity.
Control treatments reveal, in particular, that genetic information in duplicated regions sparks otherwise unlikely genome expansion.
This expansion, in turn, can promote simple adaptive traits and more coding sites --- in line with neutralist perspectives on biological complexity \citep{Lynch2000,Beslon2021,Lynch2007}.
However, we also find that genetic information within duplicated regions catalyzes complex adaptive traits, beyond effects from increased genome size alone.
Causality appears direct, as coding sites for these complex traits trace back to slip-duplicated regions.
We thus also confirm that gene duplication can facilitate otherwise unlikely complex innovation in adaptation-driven scenarios.

Figure \ref{fig:slip_mut_variants} details our experimental design.
Digital organisms were well-mixed, with population carrying capacity of 3,600.
To supply adaptive potential, rewards were provided for Lenski \textit{et al.}'s collection of nine phenotypic traits (different logical transforms on binary inputs) \citep{Lenski2003Evolutionary}.
Under this scheme, organisms producing correct output received a metabolic boost to their genetic operations.
Formally, ``substep'' complexity of each task is described as a minimum quantity of necessary NAND operations \citep{Lenski2003Evolutionary}.
\unskip\footnote{
Computer architecture theory considers NAND as a fundamental basis operation, since all other logic can be derived from compositions of NAND operations \citep{mano1997logic}.
}
The most complex function (EQU) requires five NAND operations, while the simplest (NAND and NOT) require only one.
% In other work, EQU has been shown to require at least around 20 constituent genome sites, with many possible longer ways to accomplish EQU \citep{Lenski2003Evolutionary}.

Avida organisms replicate asexually by copying their genome site-by-site.
In addition to standard point mutation, experiments
% implemented a series of mutation operators to systematically isolate aspects of gene duplication and tease apart which factors promote evolvability.
incorporated gene duplication using ``slip mutation'' events analogous to replication slippage \citep{bzymek_instability_2001}.
Under baseline conditions, these slip mutations allow arbitrary segments of program content to be duplicated or excised.
% It is also possible that ``side effects'' of gene duplication may contribute to adaptive evolution, such as effects in increasing effective mutation rate, localized clustering of sequence changes, and increases in genome length.
%Due to their inherent co-occurrence, it is not obvious how to disentangle these aspects of gene duplication in field or laboratory studies.
As overviewed in Figure \ref{fig:slip_mut_variants}, we also tested a series of partial analogs for gene duplication, designed to tease apart causality in greater detail.
Experiments began with 100-site genomes.
To control for the effects of genome length, we also included trials with longer 1,000-site genomes, near the upper extreme of 870-site genome length fixed under slip duplication (Supplementary Figure \ref{fig:ablation-site-counts}).

% \subsection{Major Results}

% We found local slip mutations that duplicate intact regions to be the most effective configuration of gene duplication in facilitating evolution on the Logic-9 task set within the Avida platform.
% In particular, we found that --- compared to control experiments with long genome sizes --- gene duplication uniquely promoted the evolution of complex adaptive traits.
% We further found that the raw material created by slip duplication plays a potentiating role in the evolution of complex traits.
% Specifically, we identified that slip-duplicated regions are significantly more likely to serve as coding sites for new traits when they first appear.
% Consistent with expectations under neofunctionalization theory, however, we did not observe potentiation effects of slip duplication on the evolution of simple traits, only traits that were facilitated by multiple building block components.

% Finally, we assessed the consequences of slip duplication on genome architecture.
% One hypothesis is that gene duplication would promote genetic brittleness by increasing genome length and accelerating growth in contingent complexity, as newly redundant genetic material specializes over time.
% Contrary to this possibility, we found that active coding sites grew as a rate similar to control experiments.
% However, we observed that slip duplications produced a significant increase in the accumulation of coding material, when active and vestigial sites are considered together.
% To understand this phenomenon, we tested the immediate effects of slip duplication on genome brittleness.
% We found that, on average, fitness-neutral slip duplications decreased, rather than increased, the number of information-baring sites in a genome.
% These results align with our observed increase in vestigial coding material in genomes.
% These brittleness-reducing effects appear to be counteracted by other factors, resulting in a similar overall trajectory of genome information content between the slip-duplication and control treatments.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% In our experiment, we used the capabilities of the digital evolution platform to compare evolutionary outcomes in terms of the complexity of traits evolved when mutational processes that duplicated genome regions were included, excluded, as well as a .

% Gene duplication events are widely recognized as a key factor in the evolution of organismal complexity.
% Mirroring broader contrasts between adaptation- versus contingency-driven hypotheses for the evolutionary origins of biological complexity, gene duplication outcomes are typically framed in terms of neo- and sub-functionalization scenarios.
% In the former, duplicated genetic material catalyzes novel functionality; in the latter, it is co-opted to elaborate existing functionality.
% Examples of both scenarios are widespread in natural history, but practical constraints have limited direct experimental investigation of the relationship between gene duplication and organismal complexity.
% Using the Avida platform for digital evolution, we show that while increased genome size can promote the emergence of simple adaptive traits, gene duplication uniquely facilitates the \textit{de novo} evolution of complex adaptive phenotypes.
% Tracing the ancestry of individual genetic sites, we find that slip duplication of a site increases its subsequent likelihood to code for novel phenotypic traits.
% We then harness the unique \textit{in silico} capabilities of our model system to compare evolutionary outcomes across degraded variants of full-fledged gene duplication.
% This ablative analysis confirms that the observed adaptive potentiation indeed arises from the duplication of existing genetic information.
% In contrast to purely neutral framings of biological complexity, our results support gene duplication events as a contributing factor in adaptive origins of complex traits.

% Concrete phenotypic effects have been directly attributed to changes in copy count, such as the short leg length characteristic of dachshunds arising from an additional copy of the FGF4 gene \citep{dachshundGeneCopyNum}.

% The prominent role of gene duplications in biological evolution has inspired incorporation of analogous mechanisms in evolution-inspired optimization algorithms \citep{Ryan:1998gm,Sawai:1999genetic,Sawai:2000comparative,Schmitt:2005bc}.
% Notably, in genetic programming, gene duplication and deletion operators have been shown to increase program evolvability and yield simpler evolved solutions \citep{Koza:1995fr}.
% In work evolving neural controllers for robots, enabling module duplication was found to increase functional specialization in network modules \citep{Calabretta:1998vh,Calabretta:2000tl}.
