\section{Results} \label{sec:results}

\subsection{Gene duplication facilitates adaptive evolution of complex traits}

\input{fig/results_panels_new.tex}

In a first set of experiments, we investigated whether gene duplication can promote \textit{de novo} evolution of adaptive phenotypic traits.
In evolutionary trials, \textit{slip-duplicate} replicates allowing duplications did evolve more adaptive traits than under baseline conditions (median 7 vs. 9 traits; two-tailed Mann-Whitney U test, $W = 562.5$, Bonferroni-adjusted $p << 0.0001$; Figure \ref{fig:results_panels:slip_duplication}).
In many cases, gene duplication drove substantial genome expansion --- inflations up to 10-fold in size, mean 546 sites (Figure \ref{fig:num-coding-sites:total}).
To test whether genome size alone could account for gene duplication's adaptive benefit, we performed additional experiments incorporating a ``long-genome'' control treatment with 1,000-site genomes.

\input{fig/adaptive-evolution-rate.tex}

The long-genome control also outperformed adaptation under 100-site baseline conditions (Supplementary Figure \ref{fig:adaptive-evolution-rate-agg}).
However, disaggregating by task complexity reveals that genome expansion alone primarily favors simple traits.
For traits requiring 3 or fewer NANDs, adaptation in the long-genome control matched or exceeded gene duplication when compared generation-by-generation (Figure \ref{fig:adaptive-evolution-rate}).
By contrast, the larger genome control failed to match duplication in promoting complex 4- and 5-NAND traits (Fisher's exact tests; 36/60 vs. 24/60, $p<0.05$ [4 components]; 10/30 vs. 2/30, $p<0.03$ [5 components]; Figure \ref{fig:adaptive-evolution-rate}).
% https://github.com/chaynes2019/AvidaGeneDupe/blob/538ede79c7301f10718ca96c8dd38782b6882632/binder/adaptive-evolution-rate.ipynb
Thus, duplication specifically advantaged complex adaptation.

% ---
% @AML: Moved to discussion
% These results align with existing findings across a wide variety of biological taxa and digital models that slip-duplication of genetic material can facilitate evolution of adaptive traits \citep{Koza:1995fr,Zhang:2003fw,Teichmann:2004cz}.
% Indeed, a striking example of gene duplication leading to new adaptations comes from the Long-Term Evolution Experiment in \textit{Escherichia coli}, where a duplication in one population broadened expression of key metabolic machinery to previously inhibitive conditions --- resulting in a 7-fold increase in population size \citep{blount_genomic_2012}.
% Among many other examples, such observations motivate longstanding connections drawn between gene duplications and adaptive traits, which date as far back as Ohno \citep{ohno1970evolution}.
% Here, systematic and controlled experimental evidence provides valuable corroboration for this understanding --- and, additionally, suggests in further detail that gene duplications may be especially important to the evolution of complex traits.
% ---

\subsection{Gene content of duplicated inserts drives adaptation}

Having observed that gene duplication can accelerate adaptive evolution beyond effects of increased genome size alone, we next sought to test whether this adaptive benefit stemmed, in specific, from genetic information propagated in duplications.
For this purpose, we tested four variants of the \textit{slip-duplication} operator, disrupting duplicated material in different ways (Figure \ref{fig:slip_mut_variants}), as well as an additional ``high mutation rate'' treatment where single-site insertion/deletion mutations were applied in lieu of duplication.

We found adaptation to strongly depend on the content of duplicated inserts.
In all cases, partial analogs of gene duplication provided no adaptive benefit --- except for \textit{slip-scramble}, which randomized sequence order within duplicated regions (two-tailed Mann-Whitney U test, $p < 0.0001$; \Cref{fig:results_panels:ablation}).
However, \textit{slip-scramble} still underperformed full-fledged \textit{slip-duplicate}, tested via 100 additional trials to avoid multiple comparison (two-tailed Mann-Whitney U test, $W = 4305$; Bonferroni-adjusted $p < 0.02$).
% p = 0.011
Sequence order of inserts notably affected genome expansion, with \textit{slip-duplicate} yielding genomes nearly twice as large as \textit{slip-scramble} (491 $\pm$ SD 223 vs. 252 $\pm$ SD 105 sites; two-tailed Mann-Whitney U test, $U = 734$, $p < 0.001$; Supplementary Figure~\ref{fig:ablation-site-counts}).

% @AML: Moved next line into discussion
% These results indicate that both the content and structure of duplicated genetic material contribute to facilitating adaptive evolution.

% As above, earlier-reported experiments within our study system concur with present work, finding sequence content and order to both also benefit adaptive evolution within changing environments \citep{lalejini2017gene}.

% \subsection{How Does Slip Duplication Facilitate Adaptive Evolution?}

% Thus far, we have established that slip-duplications of genetic material can significantly benefit adaptive evolution, and that sequence content and order both contribute to this effect.
% We next sought to explain \textit{how} the structural characteristics of slip duplications promote adaptive evolution in our study system.

% In these investigations, we evaluated three hypotheses around evolutionary dynamics of slip duplications.
% First, we explored how the facilitation of adaptive evolution by gene duplication differed between simple and complex traits promoted by gene duplication.
% Second, we tested for signatures of evolutionary potentiation by gene duplications along lineage histories.
% Finally, we examined how slip duplication influences genetic architecture with respect to brittleness and vestigial genetic material.

\subsection{Duplicated sites are potentiated for complex tasks}

% https://github.com/chaynes2019/AvidaGeneDupe/blob/4c7fa27229094adcb5bdb0b1aec541d0014b0fed/binder/hard-task-gain.ipynb
%             H-statistic       p-value
% Components
% 1              1.011820  3.144672e-01
% 2             22.787798  1.809107e-06
% 3             33.846753  5.962854e-09
% 4             15.359894  8.885441e-05
% 5              9.097744  2.559249e-03
%    Components  Prev Slip Insertion Cumulative Count      mean       std
% 0           1                                 False  0.932088  0.455109
% 1           1                                  True  1.002025  1.164197
% 2           2                                 False  0.588566  0.618479
% 3           2                                  True  1.579974  1.191681
% 4           3                                 False  0.491481  0.700256
% 5           3                                  True  1.516115  0.911430
% 6           4                                 False  0.703793  0.980105
% 7           4                                  True  1.229158  0.654042
% 8           5                                 False  0.583674  0.927917
% 9           5                                  True  1.249985  0.531423
%    Components  Prev Slip Insertion Cumulative Count  size
% 0           1                                 False    60
% 1           1                                  True    48
% 2           2                                 False    60
% 3           2                                  True    56
% 4           3                                 False    60
% 5           3                                  True    59
% 6           4                                 False    52
% 7           4                                  True    52
% 8           5                                 False    20
% 9           5                                  True    20

\input{fig/potentiation}

Thus far, we have established that content of duplicative inserts can promote evolution of novel traits, with this effect biasing toward complex traits.
% We next sought to characterize aspects of genetic architecture through which slip duplication drives adaptation.
We next sought to test whether duplicated genetic material itself exhibits elevated potential to code for novel traits.

To address the question, we measured whether coding sites for novel adaptive traits were more likely to arise in genome regions that had previously been duplicated.
High density for later coding sites would suggest that these regions are ``potentiated'' --- that is, possessing latent gene content prone to facilitate new traits \citep{Ferguson2023,Quandt2015,Blount2008}.
Figure \ref{fig:potentiation} compares the involvement of sites having undergone duplication --- versus those that had not --- in coding for novel traits.
For the simplest tasks, requiring only one NAND component, we found no potentiation by gene duplication.
% https://github.com/chaynes2019/AvidaGeneDupe/blob/2d31c0bd0cc9bdbf4bb49224859d6bd165cd36c8/binder/hard-task-gain2.ipynb
However, we found significant potentiation in duplicated regions for traits with two or more NAND components (one-sample Wilcoxon signed-rank tests; $n=56,59,52,20$ observations; Figure \ref{fig:potentiation}).
Effect sizes on probability to code for novel traits were $1.6\times$, $1.6\times$, $1.2\times$, and $1.2\times$, respectively.
Smaller effect sizes for 4- and 5-NAND tasks may in part reflect scarcity of unduplicated material at later generations (Supplementary Figure \ref{fig:potentiation-supp}).

One possible confounding factor in coding site analysis is mutational sensitivity of genome sites involved in organisms' self-replication loops.
These sites are critical to viability, with lethal outcomes when knocked out.
We found that these critical sites were indeed less likely to be involved in duplication and also less likely to be involved in coding for \textit{de novo} traits.
% Hence, a spurious correlation could be introduced.
However, after excluding fitness-critical sites from analysis, we still found generally similar potentiation signatures from duplication (Supplemental Figure \ref{fig:potentiation-supp}).

In addition to potentiating effects, gene duplications are also hypothesized to facilitate adaptation through direct effects --- for example, by altering gene product dosage \citep{kondrashov2012gene}.
% https://github.com/chaynes2019/AvidaGeneDupe/blob/61ea29d989ebc8ad83dcfee94f3fa556b81e3f78/binder/gain-mechanism.ipynb
Indeed, we observed such direct effects in our system: a substantial fraction of novel gain-of-function steps on lineages coincided with duplications --- 41 of 174, or 23.6\%.
However, in cases where duplication produced immediate gain-of-function, we still found evidence that sites coding for a new trait were more likely than chance to have been involved in earlier duplications (Supplemental Figure \ref{fig:potentiation-supp}).
% @AML, moved the line below to the discussion
% Thus, the adaptive characteristics of slip duplication observed in our system seem likely to result from a combination of potentiation and direct facilitation.

\subsection{Duplication drives accumulation of vestigial coding sites}

\input{fig/num-coding-sites}

In a final set of analyses, we broadened our scope to test how gene duplication affects overall genome architecture, focusing on robustness.
We quantified robustness by counting ``critical'' genome sites, where single-site knockouts disrupted adaptive traits or were lethal.

One conventional perspective on gene duplication is that copied genetic material can improve robustness \citep{wagner1996genetic}.
To assess this hypothesis, we assayed the impact of synthetic gene duplication mutations on evolved genomes from \textit{slip-duplicate} treatments.
For each genome, we applied one randomly-sampled duplication and measured change in critical site count.

A slight majority of sampled duplications (56\%) were fitness-neutral, producing no phenotypic changes affecting fitness.
% @AML: Moved next line to discussion
% In line with evidence that duplication-associated redundancy can boost robustness \citep{Lynch2000}, fitness-neutral slip insertions did indeed tend to reduce the number of genome sites detectable as a single point of failure for performed tasks.
% https://github.com/chaynes2019/AvidaGeneDupe/blob/9fee1f13a8d31f25d1dd799c6a26f9b7fa617738/binder/indel-effect-nulldist.ipynb
On average, we found that these fitness-neutral duplications decreased coding site count by 3\% (bootstrapped 95\% CI [6.4, 7.3] sites; Supplementary Figure \ref{fig:nulldist}). % 6.8 sites
This effect remained robust in genomes with high phenotypic complexity; for instance, neutral duplications decreased coding site count by 3\% and 5\%, respectively, in genomes with 4- and 5-NAND traits (bootstrapped 95\% CIs [8.5, 9.9] and [7.2, 9.3] sites; Supplementary Figure \ref{fig:nulldist}).
% 9.2 and 8.3 sites
Thus, short-term effects of gene duplication did, in fact, improve robustness.

Despite short-term robustness gains, however, gene duplication accrued many more critical coding sites over the long term --- 3.5$\times$ as many as baseline conditions by generation 600 (Mann-Whitney U test, $U=854$, $p << 0.0001$; Figure \ref{fig:num-coding-sites:active}). % p = 2.275468732265418e-09
At least in part, this effect likely arises from genome expansion.
By generation 600, \textit{slip-duplicate} accrued about as many active coding sites as the long-genome control (Mann-Whitney U test, $U=361$, $p=0.19$; Figure \ref{fig:num-coding-sites:active}).
% https://github.com/chaynes2019/AvidaGeneDupe/blob/binder/binder/gain-mechanism.ipynb
Despite comparable active coding sites, however, gene duplication accumulated 2.1$\times$ more vestigial coding sites (i.e., previously, but no longer, coding) compared to the long-genome control (Mann-Whitney U test, $U=630$, $p<0.01$; Figure \ref{fig:num-coding-sites:vestigial}). % p = 0.00794769551964776
Thus, gene duplication appears to drive accumulation of unconstrained coding material, potentially recruitable for later innovation.

% @AML: moved next line into discussion
% Within our study system, slip duplication appears to increase the net supply of coding material in the genome available to neutral processes, but not significantly affect accumulation of genetic brittleness.

% Indeed, in nature, large-scale genomic analyses have discovered cases where a strikingly high fraction of the genes in an organism show evidence of having arisen from gene duplications \citep{teichmann_structural_1998,Teichmann:2004cz}.
% Comparative studies have associated duplication events early in natural history with increases in genetic robustness and evolutionary innovation \citep{Wagner2008}.

% One possible explanation is that selective pressures are driving the two treatments to reach a similar number of critical sites, but slip duplication can duplicate and copy genome content in a manner that increases the copy count of previously coding sites relative to the number of critical sites.

\section{Discussion} \label{sec:discussion}

% -- Summary + key findings --
% @AML: Some of the discussion sections that I looked at (but not all) lead with a recap of authors' approach / questions.
%       - The recap can be nice for readers who jump around sections, but I'd be fine cutting it if we need the space or if folks feel that it's too redundant with the rest of the manuscript.
%      - Pretty much copied over the key findings from the cover letter
In this work, we used an \textit{in silico} model to experimentally test whether and how gene duplication can facilitate the evolution of novel adaptation and organismal complexity.
To conduct these experiments, we adapted an established framework developed by Lenski \textit{et al.} \citep{Lenski2003Evolutionary} to study adaptation across a well-defined spectrum of functional complexity.
Though necessarily limited in fully reflecting natural organisms, such digital models can nonetheless provide useful footholds to biological science \citep{Fortuna2022,Adami2000Evolution,Zaman2014Coevolution,Fortuna2017GenotypePhenotype,Gupta2016,Goldsby2012,Goldsby2014,Lenski1999,Misevic2005,Clune2008,Clune2012,Covert2013,Elena2007,Wilke2001,Hazen2007}.

Our results demonstrate that gene duplications can promote the \textit{de novo} evolution of adaptive traits.
Leveraging the unique tractability of our study system, we were able to explore causation in detail, revealing:
(1) gene duplication especially potentiated complex traits;
(2) sequence content of duplications drove adaptation; and
(3) though ultimately increasing total mutational sensitivity (despite initially buffering against it), duplicative genome expansion greatly accelerated the accumulation of neutral vestigial material.

\subsection{Gene duplication facilitates adaptive evolution of complex traits}

Adaptation by gene duplication in our study system corroborates findings across a wide variety of biological taxa and digital models \citep{Koza:1995fr,Zhang:2003fw,Teichmann:2004cz}.
Indeed, a striking example of gene duplication potentiating novel adaptation comes from the Long-Term Evolution Experiment in \textit{E. coli} \citep{lenski1991longterm,Lenski2003phenotypic}, where a duplication in one population enabled expression of a key citrate transporter in aerobic conditions --- resulting in a seven-fold increase in carrying capacity \citep{blount_genomic_2012}.
Among many other examples from natural history \citep{zhang2002adaptive,gaines2009gene,perry2007diet,hittinger2007gene,benderoth2006positive,dulai1999evolution,Ogino2015}, such observations underlie longstanding theory connecting gene duplication and adaptive traits, dating as far back as \citet{ohno1970evolution}.

On the other hand, however, recent directed evolution experiments on \textit{E. coli} notably do not find gene copy count to affect phenotypic adaptation rate \citep{mihajlovic2025direct}.
Several factors may explain this contrast, including directed evolution's shorter generational timescales, much higher mutation rates, and extreme selection strength.
Another factor may be opportunity for open-ended, multi-gene composition in our study system.

A central objective of our experimental design is not only to test whether gene duplication can facilitate adaptation, but also to identify when and how it does so.
By examining adaptive evolution across a spectrum of trait complexity, we find that gene duplication especially drives the evolution of complex traits (Figure \ref{fig:adaptive-evolution-rate}).
Comparison against a long-genome control confirms that this effect cannot be attributed to genome expansion alone.

A key aspect of Lenski \textit{et al.}'s model of phenotypic complexity is functional overlap between defined traits, meant to reflect the tendency of biological complexity to arise by modification of existing structures and functions \citep{Lenski2003Evolutionary}.
In this context, gene duplication's observed role in facilitating complex traits is broadly consistent with a ``building block'' model --- where duplication facilitates novel assemblages of existing components \citep{Jacob1977}.
Exploring how these dynamics are shaped by mechanistic complications, such as regulatory network structure \citep{CussatBlanc2019,Lalejini2021}, is an exciting direction for future work.

\subsection{Adaptation harnesses sequence information of duplications}

In order to isolate how gene duplication facilitates adaptation, we tested a suite of partial analogs for full-fledged gene duplication (Figure~\ref{fig:slip_mut_variants}).
Among these, insertions that increase genome size without propagating genetic information provided no adaptive benefit.
By contrast, both sequence duplications and shuffled duplications expedited adaptation, with the former (preserving sequence order) yielding the strongest effect (Figure~\ref{fig:results_panels}).
These results show that, rather than just side effects on genome expansion and mutation supply, adaptation can specifically harness information content of duplications --- as in classical descriptions of subfunctionalization and neofunctionalization, where gene copies gain new functionality by specializing or diverging \citep{ohno1970evolution,Lynch2000}.

In contrast to our long-genome control, no adaptive benefit arose from allowing neutral inserts (Figure~\ref{fig:results_panels}).
Despite opportunity to create increased genome length that advantages adaptation, relatively few neutral inserts accumulated.
Gene duplication induced much greater genome expansion than neutral inserts --- by a factor of nearly 4$\times$, despite equivalent site-creation potential (Supplementary Figure~\ref{fig:ablation-site-counts}).
One possible explanation why few neutral inserts fixed is selection against mutational load arising from additional genome sites.
(Indeed, in preliminary experiments we found that mutational load frequently drove genome shrinkage in founding ancestors.)
Gene duplication's strong effect on genome expansion, compared to partial analogs that do not propagate genetic information, supports theory that ``lock-in'' by dose effects or contingent interdependencies (e.g., ``complexity ratchet'' effects \citep{Liard2020,Soyer2006,Luke2011,gould1996full}) is essential to retention of new genome content \citep{Innan2010,Wagner1998,Tautz1992,Force1999,Bergthorsson2007,Conant2008,Francino2005}.

\subsection{Duplicated sites are potentiated for later adaptation}

Large-scale analyses of biological genomes report substantial content as having likely arisen by duplication \citep{teichmann_structural_1998,Teichmann:2004cz}.
In our experiments, analysis of genome history allows us to draw direct connections between gene duplication and adaptation.
Specifically, we found that (1) an outsized fraction of new adaptations arose in duplicated regions (Figure~\ref{fig:potentiation}) and (2) many gain-of-function events (41 of 174 along analyzed lineages) directly coincided with slip duplications.
Thus, adaptive outcomes appear to reflect a combination of both potentiation and immediate effects.
Lastly, echoing how gene duplication especially benefited complex adaptation, lineage analysis finds that duplicated sites potentiate only complex adaptations --- lending further credence to gene duplication as facilitating a ``building block'' mode of adaptation \citep{Jacob1977}.

\subsection{Gene duplications accumulate vestigial coding material}

Consistent with other findings that duplication-associated redundancy can boost robustness \citep{Lynch2000}, gene duplications tended to immediately reduce single points of failure detectable in knockouts (Supplemental Figure \ref{fig:nulldist}).
Over time, however, gene duplication actually led to more mutationally-sensitive coding sites --- similar in quantity to the long-genome control.
One possible explanation is that, despite duplications boosting redundancy in the short term, long-term balance between drift and mutational load drives convergence towards similar mutational sensitivity for large genomes.

Notably, gene duplication diverges from control treatments in greatly increasing net supply of vestigial coding sites no longer constrained by selection (Figure \ref{fig:num-coding-sites}).
Thus, we observe gene duplication to reshape genome architecture by expanding size, but to also produce an outsize pool of defunct coding content --- ripe to be co-opted.
Here, as throughout this study, site-by-site genome histories and robust experimental controls highlight an apparent evolutionary role of genetic information carried within duplications.
